---
title: "Take-home Exercise 2: Prototyping Modules for Shiny Application"
author: "PHAM Hung Son"
date: "Mar 02, 2026"
date-modified: "last-modified"
execute: 
  eval: true
  echo: true
  warning: false
  freeze: true
format: html
---

# 1 Introduction

Financial technology (fintech) has become a critical driver of financial inclusion and digital transformation in emerging markets. However, despite this trend, publicly available datasets integrating real transactional behavior, product usage, customer satisfaction, and digital engagement remain hard-to-find.

The **COFINFAD (Colombian Fintech Financial Analytics Dataset)** addresses this gap by providing a comprehensive 12-month dataset (January - December 2023) covering the complete active customer base of a Colombian fintech company (48,723 customers) and over 3.1 million transaction records.

## 1.1 Purpose of the EDA Module

The objective of this Exploratory Data Analysis (EDA) module is to uncover behavioral, demographic, and engagement patterns that may influence:

-   Product adoption dynamics & Digital app engagement

-   Satisfaction and Net Promoter Score (NPS)

-   Transactions Pattern & Value

-   Customer lifetime value (CLV)

## 1.2 Analytical Questions For The Module

The EDA module is structured to address the following key questions:

1.  *How are customers distributed across demographic and income segments?*

2.  *How do transaction frequency and transaction value vary across customer segments?*

3.  *What is the relationship between digital engagement (app logins, feature diversity) and transactions behavior?*

4.  *Are satisfaction and NPS aligned with behavioral indicators such as transaction volume and active products?*

# 2 Getting started

## 2.1 Installing and loading the required libraries

The R packages used in this take-home exercise are:

-   [**tidyverse**](https://www.tidyverse.org/) (i.e. readxl, tidyr, dplyr, ggplot2) for performing data science tasks such as importing, tidying, and wrangling data, as well as creating graphics;

-   [**ggthemes**](https://cran.r-project.org/web/packages/ggthemes/index.html): to use additional themes for **ggplot2**

-   [**ggridges**](https://cran.r-project.org/web/packages/ggridges/vignettes/introduction.html) for creating ridgeline plots;

-   [**ggdist**](https://mjskay.github.io/ggdist/) for visualising distributions and uncertainty;

-   [**ggstatsplot**](https://cran.r-project.org/web/packages/ggstatsplot/index.html): to create graphics with details from statistical tests; and

-   [**patchwork**](https://cran.r-project.org/web/packages/patchwork/index.html): to prepare composite figure created using **ggplot2.**

```{r}
pacman::p_load(tidyverse, lubridate, ggridges, ggdist, ggstatsplot, patchwork, plotly)
```

## 2.2 Importing data

In this exercise, we will use the The [COFINFAD](https://data.mendeley.com/datasets/mhb4zn3258/1) dataset. It is organized into two relational CSV files:

1.  **customer_data.csv:** Contains 54 customer-level attributes including demographics, product holdings, satisfaction metrics, and aggregated behavioral features (one row per customer; 48,723 rows).

2.  **transactions_data.csv:** Contains transaction records over the 12-month period (3,159,157 rows), linked via `customer_id`.

In the code chunk below, `read_csv()` is used to import the two CSV files.

```{r}
customer_data <- read_csv("data/customer_data.csv")
transactions_data <- read_csv("data/transactions_data.csv") 
```

`glimpse()` will be used to check the structure of the data.

```{r}
glimpse(customer_data)
glimpse(transactions_data)
```

## 2.3 Data Wrangling

### 2.3.1 Data Type Correction

Variables are converted into appropriate data types:

-   Categorical variables → `factor`

-   Boolean variables → `logical`

-   Date variables → `Date`

-   Numeric measures → `numeric` or `integer`

```{r}
customer_data <- customer_data %>%
  mutate(
    gender = as.factor(gender),
    income_bracket = as.factor(income_bracket),
    education_level = as.factor(education_level),
    marital_status = as.factor(marital_status),
    acquisition_channel = as.factor(acquisition_channel),
    customer_segment = as.factor(customer_segment),
    clv_segment = as.factor(clv_segment),
    savings_account = as.logical(savings_account),
    credit_card = as.logical(credit_card),
    personal_loan = as.logical(personal_loan),
    investment_account = as.logical(investment_account),
    insurance_product = as.logical(insurance_product),
    first_tx = as.Date(first_tx),
    last_tx = as.Date(last_tx),
    last_survey_date = as.Date(last_survey_date)
  )
```

```{r}
transactions_data <- transactions_data %>%
  mutate(
    date = as.Date(date),
    type = as.factor(type)
  )
```

### 2.3.2 Missing values inspection

```{r}
customer_data %>%
  summarise(across(everything(), ~sum(is.na(.))))

transactions_data %>%
  summarise(across(everything(), ~sum(is.na(.))))
```

### 2.3.3 Deriving Variables

**Age Group:** Age is grouped into interpretable bins for later filtering and segmentation analysis.

```{r}
customer_data <- customer_data %>%
  mutate(
    age = as.integer(age),  
    age_group = case_when(
      is.na(age) ~ "Unknown",
      age < 18 ~ "<18",
      age <= 25 ~ "18-25",
      age <= 35 ~ "26-35",
      age <= 45 ~ "36-45",
      age <= 55 ~ "46-55",
      age <= 65 ~ "56-65",
      TRUE ~ "65+"
    ),
    age_group = factor(
      age_group,
      levels = c("<18", "18-25", "26-35", "36-45", "46-55", "56-65", "65+", "Unknown")
    )
  )
table(customer_data$age_group, useNA = "ifany")
```

### 2.3.4 Data Validation Checks

Basic statistical summaries using `summary()` are performed to confirms data reliability before proceeding to exploratory visualization.

```{r}
summary(customer_data$age)
summary(customer_data$churn_probability)
summary(customer_data$customer_lifetime_value)
```

```{r}
write_rds(customer_data, "data/rds/customer_data.rds")
write_rds(transactions_data, "data/rds/transactions_data.rds")
```

# 3. Univariate Analysis

## 3.1 Demographic distributions of customer (age, gender, income, location)

```{r, fig.width=8, fig.height=8, out.width="100%", dpi=300}
age_group_summary <- customer_data %>%
  count(age_group) %>%
  mutate(
    prop = n / sum(n),
    label = scales::percent(prop, accuracy = 0.1)
  )
age_group_summary
```

```{r, fig.width=8, fig.height=8, out.width="100%", dpi=300}
ch1 <- ggplot(age_group_summary, aes(x = "", y = n, fill = age_group)) +
  geom_col(width = 1, color = "white") +
  coord_polar(theta = "y") +
  geom_text(aes(label = label),
            position = position_stack(vjust = 0.5),
            size = 3) +
  labs(
    title = "Customer Age Group Distribution",
    fill = "Age Group"
  ) +
  theme_void()

ch2 <- ggplot(customer_data, aes(x = gender)) +
  geom_bar(fill = "darkblue") +
  labs(
    title = "Distribution of Customer Gender",
    x = "Gender",
    y = "Number of Customers"
  ) +
  theme_minimal()

ch1 + ch2
```

```{r, fig.width=8, fig.height=8, out.width="100%", dpi=300}
ggplot(customer_data, aes(x = income_bracket)) +
  geom_bar(fill = "steelblue") +
  labs(
    title = "Distribution of Customer Income Brackets",
    x = "Income Bracket",
    y = "Number of Customers"
  ) +
  theme_minimal()
```

**Education Level**

```{r, fig.width=8, fig.height=8, out.width="100%", dpi=300}
edu_sum <- customer_data %>%
  count(education_level, sort = TRUE) %>%
  mutate(education_level = fct_inorder(education_level))

ggplot(edu_sum, aes(x = n, y = education_level)) +
  geom_segment(aes(x = 0, xend = n, yend = education_level), linewidth = 2) +
  geom_point(size = 3) +
  labs(
    title = "Educational Level Distribution",
    x = "# of Customers",
    y = "Edu Lvl"
  ) +
  theme_minimal()
```

**Martial Status**

```{r}
mar_sum <- customer_data %>%
  count(marital_status, sort = TRUE) %>%
  mutate(marital_status = fct_inorder(marital_status))

ggplot(mar_sum, aes(x = n, y = marital_status)) +
  geom_segment(aes(x = 0, xend = n, yend = marital_status), linewidth = 2) +
  geom_point(size = 3) +
  labs(
    title = "Marital Status Distribution",
    x = "Number of Customers",
    y = "Marital Status"
  ) +
  theme_minimal()
```

## 3.2 Customer Engagement Behaviour

Customer engagement is an important indicator of platform activity and potential retention.

**App Login Frequency Distribution**

```{r, fig.width=8, fig.height=8, out.width="100%", dpi=300}
ggplot(customer_data, aes(x = app_logins_frequency)) +
  stat_halfeye(
    adjust = .7,
    width = .6,
    fill = "darkgreen"
  ) +
  labs(
    title = "Distribution of Monthly App Login Frequency",
    x = "Monthly App Logins",
    y = NULL
  ) +
  theme_minimal()
```

**Satisfaction Score Distribution**

```{r, fig.width=8, fig.height=8, out.width="100%", dpi=300}
ggplot(customer_data, aes(x = satisfaction_score)) +
  geom_histogram(
    bins = 10,
    fill = "orange",
    color = "white"
  ) +
  labs(
    title = "Distribution of Customer Satisfaction Scores",
    x = "Satisfaction Score",
    y = "Frequency"
  ) +
  theme_minimal()
```

**CLV Distribution**

```{r, fig.width=8, fig.height=8, out.width="100%", dpi=300}
ggplot(customer_data, aes(x = customer_lifetime_value)) +
  geom_density(fill = "purple", alpha = 0.6) +
  scale_x_log10() +
  labs(
    title = "Distribution of Customer Lifetime Value",
    x = "Customer Lifetime Value (log scale)",
    y = "Density"
  ) +
  theme_minimal()
```

**Churn Probability Distribution**

```{r, fig.width=8, fig.height=8, out.width="100%", dpi=300}
ggplot(customer_data, aes(x = churn_probability)) +
  geom_density(fill = "red", alpha = 0.6) +
  labs(
    title = "Distribution of Customer Churn Probability",
    x = "Churn Probability",
    y = "Density"
  ) +
  theme_minimal()
```

**Product ownership distribution**

```{r, fig.width=8, fig.height=8, out.width="100%", dpi=300}
prod_long <- customer_data %>%
  select(savings_account, credit_card, personal_loan, investment_account, insurance_product) %>%
  pivot_longer(everything(), names_to = "product", values_to = "has_product") %>%
  mutate(
    has_product = if_else(has_product, "Yes", "No"),
    product = str_replace_all(product, "_", " ") %>% str_to_title()
  ) %>%
  count(product, has_product) %>%
  group_by(product) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup()

ggplot(prod_long, aes(x = product, y = prop, fill = has_product)) +
  geom_col(position = "fill") +
  coord_flip() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title = "Product Ownership Rate by Product",
    x = NULL,
    y = "Share of Customers",
    fill = "Owns Product"
  ) +
  theme_minimal()
```

**NPS distribution**

```{r, fig.width=8, fig.height=8, out.width="100%", dpi=300}
ggplot(customer_data, aes(x = nps_score, y = "")) +
  stat_slabinterval(point_interval = median_qi, .width = c(.5, .8, .95)) +
  labs(
    title = "NPS Score Distribution (Slab + Interval)",
    x = "NPS Score (0–10)",
    y = NULL
  ) +
  theme_minimal()
```

# 4. Bivariate / Multivariate Analysis

**App Login Frequency vs Churn Probability**

```{r, fig.width=8, fig.height=8, out.width="100%", dpi=300}
ggplot(customer_data, aes(x = app_logins_frequency, y = churn_probability)) +
  geom_point(alpha = 0.3, color = "steelblue") +
  geom_smooth(method = "loess", color = "red") +
  labs(
    title = "App Login Frequency vs. Churn Probability",
    x = "Monthly App Login Frequency",
    y = "Churn Probability"
  ) +
  theme_minimal()
```

**Transaction Frequency vs Satisfaction Score**

```{r, fig.width=8, fig.height=8, out.width="100%", dpi=300}
ggplot(customer_data, aes(x = tx_count, y = satisfaction_score)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "loess", color = "darkgreen") +
  labs(
    title = "Transaction Activity and Customer Satisfaction",
    x = "Total Number of Transactions",
    y = "Satisfaction Score"
  ) +
  theme_minimal()
```

**App Logins vs NPS Score**

```{r, fig.width=8, fig.height=8, out.width="100%", dpi=300}
ggplot(customer_data, aes(x = app_logins_frequency, y = nps_score)) +
  geom_jitter(alpha = 0.3, color = "purple") +
  geom_smooth(method = "loess", color = "black") +
  labs(
    title = "App Engagement and Net Promoter Score",
    x = "Monthly App Login Frequency",
    y = "NPS Score"
  ) +
  theme_minimal()
```

More comparisons to consider:

-   CLV vs active products

-   Satisfaction vs NPS

-   Transaction volume vs churn / satisfaction / NPS

# 5. Transaction Behavior

(ongoing)

-   Transaction types distribution

-   Amount distribution

-   Time trend of transactions

-   Weekend vs weekday pattern

# 6. Interactive Design for Shiny

# 7. UI Design

# Reference

-   Kam, T. S. (2023). [R for Visual Analytics](https://r4va.netlify.app/).
